---
title: "Short Salish Sea DFA"
author: "Cam Freshwater"
date: "May 3, 2019"
output: html_document
---

After preliminary runs with chinook data noticed issues with estimating states for early portion of time series. Specifically the lack of data for certain time series mean that the estimated states were predicting (potnetially) unreasonably high historical survival (these data are saved in `data/dfaFits/salishSeaOnly`). 

To explore impacts of gappy data try cropping at 1985, then refitting the DFA.

```{r loadAndClean}
listOfPackages <- c("here", "MARSS", "tidyverse", "ggplot2", "parallel", 
                    "doParallel", "foreach", "tictoc")
lapply(listOfPackages, require, character.only = TRUE)

#helper functions to fit and post-process DFA
source(here("R/functions/dfaFunctions.R"))

byDat <- read.csv(here("data/salmonData/CLEANcwtInd_age2SR_BY.csv"), 
                  stringsAsFactors = FALSE)

#focus on subset of BC pops for initial analyses
byDatTrim <- byDat %>% 
  filter(region %in% c("LFR", "MFR", "UFR", "ECVI", "SPGSD", "NPGSD"))%>% 
  group_by(stock) %>% 
  mutate(survZ = as.numeric(scale(surv))) %>%
  ungroup(stock) %>% 
  arrange(region) %>% 
  mutate(stock = factor(stock, unique(stock))) %>% 
  select(-stockName, -jurisdiction, -lat, -long) %>% 
  filter(BY > 1984)

byMatZ <- byDatTrim %>%
  select(BY, stock, survZ) %>% 
  spread(key = stock, value = survZ) %>% 
  select(-BY) %>% 
  as.matrix() %>% 
  t()
broodYrs <- unique(byDatTrim$BY)
colnames(byMatZ) <- broodYrs

nStks <- nrow(byMatZ)
nYrs <- ncol(byMatZ)
stkID <- rownames(byMatZ)
```

Now fit the DFA. Focus on diagonal and unequal observation variances only given evidence from initial AICc.

```{r fitDFA, eval=FALSE, include=FALSE}
inMList <- list(1, 2, 3, 4, 5, 6, 7)
subDir <- "salishSeaOnly_short" #for data cropped at 1985
inRSeq <- c("diagonal and unequal", "equalvarcov")

for (i in seq_along(inRSeq)) {
  inRDum <- inRSeq[i]
  Ncores <- detectCores()
  cl <- makeCluster(Ncores - 2) #save two cores
  registerDoParallel(cl)
  clusterEvalQ(cl, c(library(MARSS), library(here), library(Rcpp),
                     library(RcppArmadillo)))
  clusterExport(cl, c("byMatZ", "inMList", "fitDFA", "subDir", "inRDum"),
                envir=environment())
  tic("run in parallel")
  parLapply(cl, inMList, function(x) {
    fitDFA(byMatZ, inR = inRDum, inM = x, maxIteration = 1000, 
           subDirName = subDir)
  })
  stopCluster(cl) #end cluster
  toc()
}
```

Strong support (somewhat surprisingly strong) for including only two trends. 

```{r aic}
dfaSummary <- getTopDFA(subDir)
dfaSummary[[2]] #prints AIC table

mod1 <- dfaSummary[[1]]
```

Most stocks load similarly onto trend 1 and 2.

```{r factorLoadings}
estZ <- coef(mod1, type = "matrix")$Z
#retrieve rotated matrix
invH <- if (ncol(estZ) > 1) {
  varimax(estZ)$rotmat
} else if (ncol(estZ) == 1) {
  1
}

## Factor loadings
#rotate factor loadings
rotZ <- rotateLoadings(zIn = estZ, H = invH, stkNames = stkID, 
                      survDat = byDatTrim)

ggplot(rotZ, aes(x = stock, y = loading, fill = region)) +
  geom_col() +
  theme_sleekX(axisSize = 9, legendSize = 0.7) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  facet_wrap(~trend)
```

```{r trends}
rotTrends <- rotateTrends(modIn = mod1, H = invH)

ggplot(rotTrends, aes(x = year, y = est)) +
  geom_line() +
  theme_sleekX() +
  geom_hline(yintercept = 0, colour = "red") +
  facet_wrap(~trend)
```

First fitted states based on `broom::augment` estimates.

```{r fittedStates1}
modOutCI <- broom::augment(mod1, interval = "confidence") %>% 
  dplyr::rename(stock = .rownames) %>% 
  inner_join(byDatTrim %>% select(stock, region), by = "stock") %>% 
  distinct() %>% 
  mutate(year = t + 1984, #add so that year is correct 
         stock = as.factor(stock)) %>% 
  arrange(region) 

ggplot(modOutCI) +
  geom_line(aes(x = year, y = .fitted)) +
  geom_point(aes(x = year, y = y, colour = region)) +
  geom_ribbon(aes(x = year, ymin = .conf.low, ymax = .conf.up), linetype = 2, 
              alpha = 0.2) +
  facet_wrap(~stock)
```

Second fitted states based on old school function by Mark Scheuerrell.

```{r fittedStates2}
dum <- get_DFA_fits(mod1)
namesIn <- names(dum)
dum2 <- lapply(seq_along(dum), function (x) gatherList(dum[[x]], yrs = broodYrs, 
                                                       stks = stkID, 
                                                       names = namesIn[x])) 
modOutCI2 <- do.call(rbind, dum2) %>% 
  spread(key = estimate, value = value) %>% 
  inner_join(byDatTrim %>% select(stock, region), by = "stock") %>% 
  distinct() %>% 
  mutate(stock = as.factor(stock)) %>% 
  arrange(region) %>% 
  filter(year > 1985)
  
ggplot(modOutCI2) +
  geom_line(aes(x = year, y = ex)) +
  geom_point(aes(x = modOutCI$year, y = modOutCI$y, 
                 colour = modOutCI$region)) +
  geom_ribbon(aes(x = year, ymin = lo, ymax = up), linetype = 2, 
              alpha = 0.2) +
  facet_wrap(~stock)
```